---
title: "GDP and life expectancy globally between countries over the years"
author: "Studentnames and studentnumbers here"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document:
    df_print: paged
bibliography: references.bib
csl: apa.csl
---

# Set-up your environment

```{r package_install, include=FALSE}
#install.packages("tidyverse")
#install.packages("WDI")
#install.packages("tmap")
require(tidyverse)
require(rmarkdown)
require(yaml)
library(tidyr)
library(writexl)
library(readxl)
```

# Title Page
A new sentence
Include your names

Include the tutorial group number
3.6
Include your tutorial lecturer's name
Chantal Schouwenaar
# Part 1 - Identify a Social Problem
Use APA referencing throughout your document. [Here's a link to some explanation.](https://www.mendeley.com/guides/apa-citation-guide/)
## 1.1 Describe the Social Problem

Include the following:

-   Why is this relevant? [@smith2020]

-   ...
# Part 2 - Data Sourcing

## 2.1 Load in the data

Preferably from a URL, but if not, make sure to download the data and store it in a shared location that you can load the data in from. Do not store the data in a folder you include in the Github repository!

```{r loading_data}
#dataset <- midwest

library(readr)
#Gezondheid_leefstijl_zorggebruik_en_aanbod_doodsoorzaken_kerncijfers_03062026_132800 <- #read_delim("~/Downloads/Gezondheid__leefstijl__zorggebruik_en__aanbod__doodsoorzaken__kerncijfers_03062026_132800.csv", 
#    delim = ";", escape_double = FALSE, trim_ws = TRUE, 
#    skip = 4)
#View(Gezondheid_leefstijl_zorggebruik_en_aanbod_doodsoorzaken_kerncijfers_03062026_132800)

#write_csv(Gezondheid_leefstijl_zorggebruik_en_aanbod_doodsoorzaken_kerncijfers_03062026_132800, "data/gezondheid.csv")

a_fun_name <- read_delim("data/gezondheid.csv")

a_fun_name$...1 <- NULL


write_csv(a_fun_name, "data/gezondheid_changed.csv")
```

midwest is an example dataset included in the tidyverse package

## 2.2 Provide a short summary of the dataset(s)

```{r}
head(dataset)
```

In this case we see 28 variables, but we miss some information on what units they are in. We also don't know anything about the year/moment in which this data has been captured.


```{r}


library(readr)
API_NY_GDP_PCAP_PP_KD_DS2_en_csv_v2_1700 <- read_csv("API_NY/API_NY.GDP.PCAP.PP.KD_DS2_en_csv_v2_1700.csv", 
    skip = 3)
View(API_NY_GDP_PCAP_PP_KD_DS2_en_csv_v2_1700)

#change data name:
library(readr)

gdp_raw <- read_csv(
  "API_NY/API_NY.GDP.PCAP.PP.KD_DS2_en_csv_v2_1700.csv",
  skip = 3
)

View(gdp_raw)

# cleaning the data for 10 countries and 5 years
library(dplyr)

gdp_selected <- gdp_raw %>%
  select(
    country = `Country Name`,
    country_code = `Country Code`,
    `2019`,
    `2020`,
    `2021`,
    `2022`,
    `2023`
  )

View(gdp_selected)

#Filter selected countries
selected_countries <- c(
  "USA",
  "NLD",
  "DEU",
  "JPN",
  "CHN",
  "IND",
  "BRA",
  "ZAF",
  "NGA",
  "AUS"
)

gdp_10countries <- gdp_selected %>%
  filter(country_code %in% selected_countries)

View(gdp_10countries)


saveRDS(gdp_10countries, "gdp_10countries.rds")


```

Please use a separate 'R block' of code for each type of cleaning. So, e.g. one for missing values, a new one for removing unnecessary variables etc.



```{r}

library(readr)
Life_Expectancy_raw_data <- read_csv("API_SP/API_SP.DYN.LE00.IN_DS2_en_csv_v2_265357.csv")

#changing the name of data life expectancy

life_raw <- read_csv("API_SP/API_SP.DYN.LE00.IN_DS2_en_csv_v2_265357.csv", skip = 3)

View(life_raw)

#installing package

install.packages("tidyverse")

#filtering 5 years for life expectancy

library(tidyverse)
library(dplyr)

life_select <- life_raw %>%
  select(
    country = "Country Name",
    country_code = "Country Code",
    "2019",
    "2020",
    "2021",
    "2022",
    "2023")

View(life_select)

#filtering 10 countries for life expectancy

selected_countries <- c(
  "USA",
  "NLD",
  "DEU",
  "JPN",
  "CHN",
  "IND",
  "BRA",
  "ZAF",
  "NGA",
  "AUS"
)

life_expectancy_10countries <- life_select %>%
  filter(country_code %in% selected_countries)

View(life_expectancy_10countries)

saveRDS(life_expectancy_10countries, "life_expectancy.rds")

```






```{r}
#merge the data and make the plots
#install.packages("tidyverse")
library(tidyverse)

gdp_10countries <- readRDS("gdp_10countries.rds")

life_expectancy_10countries <- readRDS("life_expectancy.rds")
View(gdp_10countries)

View(life_expectancy_10countries)

#Convert GDP data to long format
gdp_long <- gdp_10countries %>%
  pivot_longer(
    cols = `2019`:`2023`,
    names_to = "year",
    values_to = "gdp_per_capita"
  )
View(gdp_long)

#Convert life expectancy data to long format
life_long <- life_expectancy_10countries %>%
  pivot_longer(
    cols = `2019`:`2023`,
    names_to = "year",
    values_to = "life_expectancy"
  )
View(life_long)

#combine 2 data sets
combined_data <- gdp_long %>%
  left_join(
    life_long,
    by = c("country", "country_code", "year")
  ) %>%
  mutate(
    year = as.numeric(year)
  )
view(combined_data)

saveRDS(
  combined_data,
  "combined_data.rds"
)
```


```{r}
#plotting average of 10 countries life expectancy
library(ggplot2)
library(dplyr)

# Calculate average life expectancy per year
avg_life <- combined_data %>%
  group_by(year) %>%
  summarise(
    avg_life_expectancy = mean(life_expectancy, na.rm = TRUE)
  )
# Plot
ggplot(avg_life, aes(x = year, y = avg_life_expectancy)) +
  geom_line(color = "darkgreen", linewidth = 1.2) +
  geom_point(size = 3) +
  theme_minimal() +
  labs(
    title = "Average Life Expectancy of 10 Countries (2019–2023)",
    x = "Year",
    y = "Average Life Expectancy"
  )
ggsave("average_life_expectancy.png", width = 8, height = 5, dpi = 300)
```





#makeing the scatter plot

```

## 3.2 Generate necessary variables

Variable 1

```{r gen_var1}

```

Variable 2

```{r gen_var2}

```

## 3.3 Visualize temporal variation

```{r}

```

## 3.4 Visualize spatial variation

```{r visualise_map}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.5 Visualize sub-population variation

What is the poverty rate by state?

```{r visualise_subpopulations}
dataset$inmetro <- dataset$inmetro %>% as.factor()
# Boxplot of poverty rate by state using the 'midwest' dataset
ggplot(dataset, aes(x = inmetro, y = percadultpoverty)) +
  geom_boxplot() +
  labs(
    title = "Distribution of Poverty Rates by Metropolitan status (Midwest counties)",
    x = "Metropolitan Area",
    y = "Poverty Rate of Adults (%)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right"
  )
```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.6 Event analysis

Analyze the relationship between two variables.

```{r analysis}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

# Part 4 - Discussion

## 4.1 Discuss your findings

# Part 5 - Reproducibility

## 5.1 Github repository link

Provide the link to your PUBLIC repository here: ...

## 5.2 Reference list

Use APA referencing throughout your document.
