---
title: "GDP and life expectancy globally between countries over the years"
author: 
date: "`r Sys.Date()`"
output: pdf_document
bibliography: references.bib
csl: apa.csl

---

Thea von Sydow 2882066


# Set-up your environment

```{r package_install, include=FALSE}
install.packages("tidyverse")
install.packages("WDI")
install.packages("tmap")
require(tidyverse)
require(rmarkdown)
require(yaml)
library(tidyr)
library(writexl)
library(readxl)
```

# Title Page
A new sentence
Include your names

Tutorial group: 3.6

Tutorial lecturer: Chantal Schouwenaar

# Part 1 - Identify a Social Problem

## 1.1 Describe the Social Problem

Differences in life expectancy between countries are one of the clearest forms of 
global inequality. A child born in a high-income country can expect to live around 
fifteen to twenty years longer than a child born in a low-income country, and this 
gap has persisted for decades. The classic explanation links population health to 
national income: (Preston, 1975) showed that life expectancy rises steeply with GDP 
per capita at low income levels and flattens among richer countries, which implies 
that economic growth buys the most health where countries are poorest (Deaton, 2003)]. 
A key lever for policy is therefore knowing how tightly health follows income, 
because that determines whether health gaps should be fought with economic 
development, with health spending, or both.

What is less well established is how this relationship behaves during a global 
shock. The COVID-19 pandemic caused the first worldwide drop in life expectancy 
in decades, and the losses were very unequal between countries (Schöley, 2022), 
yet income alone did not fully predict who lost the most. In this report we 
quantify the relationship between GDP per capita and life expectancy for ten 
countries across all income levels between 2019 and 2023, and we examine how 
life expectancy changed around the 2020 outbreak of the COVID-19 pandemic.

# Part 2 - Data Sourcing<

## 2.1 Load in the data

Preferably from a URL, but if not, make sure to download the data and store it in a shared location that you can load the data in from. Do not store the data in a folder you include in the Github repository!

```{r loading_data}
#dataset <- midwest

library(readr)
#Gezondheid_leefstijl_zorggebruik_en_aanbod_doodsoorzaken_kerncijfers_03062026_132800 <- #read_delim("~/Downloads/Gezondheid__leefstijl__zorggebruik_en__aanbod__doodsoorzaken__kerncijfers_03062026_132800.csv", 
#    delim = ";", escape_double = FALSE, trim_ws = TRUE, 
#    skip = 4)
#View(Gezondheid_leefstijl_zorggebruik_en_aanbod_doodsoorzaken_kerncijfers_03062026_132800)

#write_csv(Gezondheid_leefstijl_zorggebruik_en_aanbod_doodsoorzaken_kerncijfers_03062026_132800, "data/gezondheid.csv")

a_fun_name <- read_delim("data/gezondheid.csv")

a_fun_name$...1 <- NULL


write_csv(a_fun_name, "data/gezondheid_changed.csv")
```

midwest is an example dataset included in the tidyverse package

## 2.2 Provide a short summary of the dataset(s)

```{r}
head(dataset)
```

In this case we see 28 variables, but we miss some information on what units they are in. We also don't know anything about the year/moment in which this data has been captured.


```{r}


library(readr)
API_NY_GDP_PCAP_PP_KD_DS2_en_csv_v2_1700 <- read_csv("API_NY/API_NY.GDP.PCAP.PP.KD_DS2_en_csv_v2_1700.csv", 
    skip = 3)
View(API_NY_GDP_PCAP_PP_KD_DS2_en_csv_v2_1700)

#change data name:
library(readr)

gdp_raw <- read_csv(
  "API_NY/API_NY.GDP.PCAP.PP.KD_DS2_en_csv_v2_1700.csv",
  skip = 3
)

View(gdp_raw)

# cleaning the data for 10 countries and 5 years
library(dplyr)

gdp_selected <- gdp_raw %>%
  select(
    country = `Country Name`,
    country_code = `Country Code`,
    `2019`,
    `2020`,
    `2021`,
    `2022`,
    `2023`
  )

View(gdp_selected)

#Filter selected countries
selected_countries <- c(
  "USA",
  "NLD",
  "DEU",
  "JPN",
  "CHN",
  "IND",
  "BRA",
  "ZAF",
  "NGA",
  "AUS"
)

gdp_10countries <- gdp_selected %>%
  filter(country_code %in% selected_countries)

View(gdp_10countries)


saveRDS(gdp_10countries, "gdp_10countries.rds")


```

Please use a separate 'R block' of code for each type of cleaning. So, e.g. one for missing values, a new one for removing unnecessary variables etc.



```{r}

library(readr)
Life_Expectancy_raw_data <- read_csv("API_SP/API_SP.DYN.LE00.IN_DS2_en_csv_v2_265357.csv")

#changing the name of data life expectancy

life_raw <- read_csv("API_SP/API_SP.DYN.LE00.IN_DS2_en_csv_v2_265357.csv", skip = 3)

View(life_raw)

#installing package

install.packages("tidyverse")

#filtering 5 years for life expectancy

library(tidyverse)
library(dplyr)

life_select <- life_raw %>%
  select(
    country = "Country Name",
    country_code = "Country Code",
    "2019",
    "2020",
    "2021",
    "2022",
    "2023")

View(life_select)

#filtering 10 countries for life expectancy

selected_countries <- c(
  "USA",
  "NLD",
  "DEU",
  "JPN",
  "CHN",
  "IND",
  "BRA",
  "ZAF",
  "NGA",
  "AUS"
)

life_expectancy_10countries <- life_select %>%
  filter(country_code %in% selected_countries)

View(life_expectancy_10countries)

saveRDS(life_expectancy_10countries, "life_expectancy.rds")

```






```{r}
#merge the data and make the plots
#install.packages("tidyverse")
library(tidyverse)

gdp_10countries <- readRDS("gdp_10countries.rds")

life_expectancy_10countries <- readRDS("life_expectancy.rds")
View(gdp_10countries)

View(life_expectancy_10countries)

#Convert GDP data to long format
gdp_long <- gdp_10countries %>%
  pivot_longer(
    cols = `2019`:`2023`,
    names_to = "year",
    values_to = "gdp_per_capita"
  )
View(gdp_long)

#Convert life expectancy data to long format
life_long <- life_expectancy_10countries %>%
  pivot_longer(
    cols = `2019`:`2023`,
    names_to = "year",
    values_to = "life_expectancy"
  )
View(life_long)

#combine 2 data sets
combined_data <- gdp_long %>%
  left_join(
    life_long,
    by = c("country", "country_code", "year")
  ) %>%
  mutate(
    year = as.numeric(year)
  )
view(combined_data)

saveRDS(
  combined_data,
  "combined_data.rds"
)
```


```{r}
#plotting average of 10 countries life expectancy
library(ggplot2)
library(dplyr)

# Calculate average life expectancy per year
avg_life <- combined_data %>%
  group_by(year) %>%
  summarise(
    avg_life_expectancy = mean(life_expectancy, na.rm = TRUE)
  )
# Plot
ggplot(avg_life, aes(x = year, y = avg_life_expectancy)) +
  geom_line(color = "darkgreen", linewidth = 1.2) +
  geom_point(size = 3) +
  theme_minimal() +
  labs(
    title = "Average Life Expectancy of 10 Countries (2019–2023)",
    x = "Year",
    y = "Average Life Expectancy"
  )
ggsave("average_life_expectancy.png", width = 8, height = 5, dpi = 300)
```





#makeing the scatter plot

```

## 3.2 Generate necessary variables

Variable 1

```{r gen_var1}

```

Variable 2

```{r gen_var2}

```

## 3.3 Visualize temporal variation

```{r}

```

## 3.4 Visualize spatial variation

```{r visualise_map}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.5 Visualize sub-population variation

What is the poverty rate by state?

```{r visualise_subpopulations}
dataset$inmetro <- dataset$inmetro %>% as.factor()
# Boxplot of poverty rate by state using the 'midwest' dataset
ggplot(dataset, aes(x = inmetro, y = percadultpoverty)) +
  geom_boxplot() +
  labs(
    title = "Distribution of Poverty Rates by Metropolitan status (Midwest counties)",
    x = "Metropolitan Area",
    y = "Poverty Rate of Adults (%)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right"
  )
```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.6 Event analysis

Analyze the relationship between two variables.

```{r analysis}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

# Part 4 - Discussion

## 4.1 Discuss your findings

# Part 5 - Reproducibility

## 5.1 Github repository link

Provide the link to your PUBLIC repository here: ...

## 5.2 Reference list

Cutler, D., Deaton, A., & Lleras-Muney, A. (2006). The determinants of mortality. Journal of Economic Perspectives, 20(3), 97–120. https://doi.org/10.1257/jep.20.3.97

Deaton, A. (2003). Health, inequality, and economic development. Journal of Economic Literature, 41(1), 113–158. https://doi.org/10.1257/002205103321544710

Preston, S. H. (1975). The changing relation between mortality and level of economic development. Population Studies, 29(2), 231–248. https://doi.org/10.1080/00324728.1975.10410201

Schöley, J., Aburto, J. M., Kashnitsky, I., Kniffka, M. S., Zhang, L., Jaadla, H., Dowd, J. B., & Kashyap, R. (2022). Life expectancy changes since COVID-19. Nature Human Behaviour, 6, 1649–1659. https://doi.org/10.1038/s41562-022-01450-3

World Health Organization. (2023). World health statistics 2023: Monitoring health for the SDGs. https://www.who.int/publications/i/item/9789240074323
